---
title: "Data Wrangling"
author: 
- name: "Adrien Osakwe"
  affiliation: Workshop Lead
- name: Larisa M. Soto
  affiliation: Original Material
date:  "`r format(Sys.Date(), '%B %d, %Y')`"
output:
   rmdformats::html_clean:
    toc: true
    thumbnails: false
    floating: true
    highlight: kate
    use_bookdown: true
---

# Data manipulation with dplyr

**Learning objectives**

-   Become familiar with the `dplyr` syntax
-   Create pipes with the operator %\>%
-   Perform operations on data frames using dplyr and tidyr functions
-   Implement functions from other external packages

There are several packages that allow for more sophisticated processing operations to be done faster. We will take a look at some functions from one of them. I encourage you to look into `plyr` and `tidyr` after this workshop.

For this section we will use data from the `gapminder` package.

```{r}
library(dplyr)
library(gapminder)
dim(gapminder)
```

View the data frame

```{r}
View(gapminder)
```

```{r}
summary(gapminder$lifeExp)
```

## Manipulation with `dplyr`

We often need to select certain observations (rows) or variables (columns), or group the data by certain variable(s) to calculate some summary statistics. Although these operations can be done using base R functions, they require the creation of multiple intermediate objects and a lot of code repetition. There are two packages that provide functions to streamline common operations on tabular data and make the code look nicer and cleaner.

These packages are part of a broader family called `tidyverse`, for more information you can visit <https://www.tidyverse.org/>.

We will cover 5 of the most commonly used functions and combine them using pipes (`%>%`):

1\. `select()` - used to extract data

2\. `filter()` - to filter entries using logical vectors

3\. `group_by()` - to solve the split-apply-combine problem

4\. `summarize()` - to obtain summary statistics

5\. `mutate()` - to create new columns

```{r}
library(tidyr)
```

### Introducing pipes

```{r}
head(gapminder)
gapminder %>%
  head()
gapminder %>%
  tail()
```

```{r}
x <- 100
log10(sqrt(x))

x %>%
  sqrt() %>%
  log10()
```

### Using `select()`

To subset a data frame

```{r}
dplyr::select(.data = gapminder, 
       year, country, gdpPercap) %>%
  head()

head(gapminder[,c("year","country","gdpPercap")])
```

To remove columns

```{r}
dplyr::select(.data = gapminder, 
       -continent) %>%
      head()
```

```{r}
gapminder %>% 
  dplyr::select(year, country, gdpPercap) %>%
  head()
```

### Using `filter()`

Include **only European countries** and select the following columns; year, country and gdpPercap

```{r}
gapminder %>%
    dplyr::filter(continent == "Europe") %>%
    dplyr::select(year, country, gdpPercap) %>%
    head()
```

Using multiple filters at once

```{r}
gapminder %>%
  dplyr::filter(continent == "Europe", year == 2007) %>%
  dplyr::select(country, lifeExp)
```

Extract unique entries

```{r}
gapminder %>%
  dplyr::select(country, continent) %>%
  dplyr::distinct()
```

Order according to a column

```{r}
gapminder %>%
  dplyr::select(country, continent,year,pop) %>%
  dplyr::arrange(desc(pop)) %>%
  head()
```

### Using `group_by()`

It internally groups observations based on the specified variable(s)

```{r}
str(gapminder)
```

```{r}
str(gapminder %>% dplyr::group_by(continent))
```

### Using `summarize()`

```{r}
gdp_c <- gapminder %>%
          dplyr::group_by(continent) %>%
          dplyr::summarize(mean_gdpPercap = mean(gdpPercap))
gdp_c
```

Combine multiple summary statistics

```{r}
gapminder %>%
    dplyr::group_by(continent) %>%
    dplyr::summarize(mean_le = mean(lifeExp),
                      min_le = min(lifeExp),
                      max_le = max(lifeExp),
                      se_le = sd(lifeExp)/sqrt(dplyr::n()))
```

### Using `mutate()`

```{r}
gapminder %>%
  dplyr::mutate(gdp_billion = gdpPercap*pop/10^9)
```

```{r}
new_df <- gapminder %>% 
  mutate(abc = ifelse(lifeExp >= 60, 'A','B')) %>%
  group_by(abc) %>%
  summarize(mean_abc_gdp = mean(gdpPercap))

new_df
```

### Putting them all together

```{r}
gdp_pop_ext <- gapminder %>%
                dplyr::mutate(gdp_billion = gdpPercap*pop/10^9) %>%
                dplyr::group_by(continent,year) %>%
                dplyr::summarize(mean_gdpPercap = mean(gdpPercap),
                                 sd_gdpPercap = sd(gdpPercap),
                                 mean_pop = mean(pop),
                                 sd_pop = sd(pop),
                                 mean_gdp_billion = mean(gdp_billion),
                                 sd_gdp_billion = sd(gdp_billion)) 

gdp_pop_ext
```

## Hands-on advanced data manipulation

1.  Write one command (it can span multiple lines) using pipes that will output a data frame that has only the columns `lifeExp`, `country` and `year` for the records before the year 1977 from African countries, but not for other Continents.

    ```{r}

    ```

    \

2.  Calculate the average life expectancy per country. Which country has the longest average life expectancy and which one the shortest average life expectancy?

    ```{r}

    ```

    \

3.  In the previous hands-on you discovered that all the entries from 2007 are actually from 2008. Write a command to edit the data accordingly using pipes. In the same command filter only the entries from 2008 to verify the change.

    ```{r}

    ```

## Data Cleaning

First thing we could do is to drop any row that includes a missing entry. This can be done with `drop_na()` from the package `tidyr` (this package is loaded with `tidyverse`).

```{r}
# Creating Missing data
noisy_gap <- gapminder
missing_idx <- sample(1:nrow(noisy_gap),
                      350)

#Adding Noisy entries
noisy_gap[missing_idx,"lifeExp"] <- -20
print(nrow(noisy_gap))

#Modify incorrect entries to mark as NA
noisy_gap <- noisy_gap %>%
  mutate(lifeExp = ifelse(lifeExp > 0,lifeExp,NA))


noisy_gap %>% 
  drop_na()
```

It is often the case that removing rows with missing entries causes too much data loss. To this end, it is preferable to fill the entries with **imputed** values. A simple way of doing this is with the `fill()` from `tidyr`. By default, the entries will be filled with the same value as the next non-missing entry below. This means that the **order** of the entries is important. You can consider making different assumptions to decide how they should be ordered. In our case, we see that life-expectancy has a reasonable correlation with GDP per-capita so we can consider ordering the data by gdpPercap to fill in missing lifeExp entries.

```{r }
# Checking correlation bet
noisy_gap %>% 
  drop_na() %>%
  select(lifeExp,gdpPercap) %>%
  cor()

##Keeping track of original order
noisy_gap$Index <- 1:nrow(noisy_gap)

noisy_gap <- noisy_gap %>% 
  dplyr::arrange(desc(gdpPercap)) %>%
  fill(lifeExp,.direction = 'downup')

#Realigning Datasets
noisy_gap <- noisy_gap[order(noisy_gap$Index),]

#Checking correlation between true and imputed
overall_correlation <- cor(gapminder$lifeExp, noisy_gap$lifeExp)
missing_idx_correlation <- cor(gapminder$lifeExp[missing_idx], noisy_gap$lifeExp[missing_idx])

# Print correlations
cat("Overall correlation:", overall_correlation, "\n")
cat("Correlation for missing indices:", missing_idx_correlation, "\n")
```

Fill entries with the sample average/median

```{r}
noisy_gap[missing_idx,"lifeExp"] <- -20
noisy_gap <- noisy_gap %>%
  mutate(lifeExp = ifelse(lifeExp > 0,lifeExp,NA))

med_le <- median(noisy_gap$lifeExp,na.rm = T)
avg_le <- mean(noisy_gap$lifeExp, na.rm = T)

med_cor <- cor(gapminder$lifeExp, ifelse(is.na(noisy_gap$lifeExp),med_le,noisy_gap$lifeExp))
avg_cor <- cor(gapminder$lifeExp, ifelse(is.na(noisy_gap$lifeExp),avg_le,noisy_gap$lifeExp))

cat("Overall correlation with median:", med_cor, "\n")
cat("Overall correlation with mean:", avg_cor, "\n")

```

Repeat above but using Country-specific Averages

```{r}
noisy_gap[missing_idx,"lifeExp"] <- -20
noisy_gap <- noisy_gap %>%
  mutate(lifeExp = ifelse(lifeExp > 0,lifeExp,NA))

country_avg <- noisy_gap %>%
  group_by(country) %>%
  summarize(avg_lifeExp = mean(lifeExp, na.rm = TRUE), .groups = "drop")

noisy_gap <- noisy_gap %>%
  left_join(country_avg, by = "country") %>%
  mutate(lifeExp = ifelse(is.na(lifeExp), avg_lifeExp, lifeExp)) %>%
  select(-avg_lifeExp)

# Check correlation after filling with country-specific averages
country_avg_cor <- cor(gapminder$lifeExp, noisy_gap$lifeExp)
country_avg_miss_cor <- cor(gapminder$lifeExp[missing_idx], noisy_gap$lifeExp[missing_idx])
cat("Overall correlation with Country-specific means:", country_avg_cor, "\n")
cat("Overall correlation with Country-specific means for missing data:", country_avg_miss_cor, "\n")

```

We can see that for both the overall dataset and the missing entries, the imputations are a lot more accurate when split by condition. We could also consider imputing by a combination of factors by training a model to predict the missing entries.

# References

-   [Base R Cheat Sheet](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)\
-   [Google's R Style Guide](https://google.github.io/styleguide/Rguide.html)
