---
title: "Statistical Tests"
author: 
- name: "Adrien Osakwe"
  affiliation: Workshop Lead
- name: "Alex Diaz-Papkovich"
  affiliation: Original Material
date:  "`r format(Sys.Date(), '%B %d, %Y')`"
output:
   rmdformats::html_clean:
    toc: true
    thumbnails: false
    floating: true
    highlight: kate
    use_bookdown: true
---

## Hypothesis testing

### The t-test

```{r A basic example of the t-test}
set.seed(20221011) # Seed for reproducibility
n <- 100 # Sample size

mu1 <- 0 # Mean of sample 1
mu2 <- 0.1 # Mean of sample 2

sd1 <- 1 # Standard deviation 1
sd2 <- 1 # Standard deviation 2

# Generate two samples of of size n of normally-distributed data.
# Sample x1 has a mean of mu1 and standard deviation sd1
# Sample x2 has a mean of mu2 and standard deviation sd2
x1 <- rnorm(n, mu1, sd1)
x2 <- rnorm(n, mu2, sd2)

```

```{r}
x_df <- data.frame(Weight = c(x1,x2),
                   Group = c(rep('H',n),rep('I',n)))

t.test(x_df$Weight[x_df$Group =='H'],x_df$Weight[x_df$Group =='I'])
```

There are many ways to look at our data. Two examples are: 1. Density plots 2. Box plots

```{r Looking at your data with a density plot}
# First look at your data
# We draw density plots to see if our data are different
# Overlay the two plots
plot(density(x1), xlim=c(-5,5), ylim=c(0,0.5), col = 1, lty = 2, main = "Comparing two population means")
abline(v=mu1, col=1, lty = 2)
lines(density(x2), col = 3, lty = 4)
abline(v=mu2, col=3, lty = 4)
```

```{r Looking at your data with box plots}
# The boxplot() function requires a data frame
# We'll put our data into a data frame with one column representing the random data and the other representing the group label
df <- data.frame(c(x1,x2),
                    c(rep(1,length(x1)), rep(2,length(x2))))
colnames(df) <- c("value","group")
boxplot(value ~ group, data = df)
```

```{r Carrying out a t-test}
# Now run a t-test comparing the two means. 
# Do you reject your null hypothesis?
# How does it change if you change n or mu1 or mu2?

# The t.test() function has several parameters
# If we put in two vectors, we can compare them
#sample size of x1 now is twice x2.
t.test(rep(x1,2), x2) 
```

```{r More t-test parameters}
# We can get more degrees of freedom with an equal-variance t-test (though it doesn't really matter once n is around 30+)
# This is a logical variable that returns whether the standard deviations of our two samples are equal
# If they are, it returns true
# Try putting sd1==sd2 into the console and seeing what it returns!
ve <- sd1 == sd2 #which version if t-test we are doing
#if equal, pooled t-test, else, Welch's t-test (default).

t.test(x1,x2,var.equal = ve)
```

```{r One-sided t-test}
# For a one-sided t-test of whether a value is greater than zero, just use one parameter:
t.test(x2)
t.test(x2, alternative = c("greater")) # or specify "greater"
```

We can retrieve relevant values by storing them in an R object and using `$` to extract them.

```{r Storing and retrieving results}
t_test_results <- t.test(x1, x2)

names(t_test_results) # This will tell you what is store in the t_test_results object
t_test_results$data.name
```

### t-test exercises

As an exercise, extract the t-statistic and p-value you calculated above using `$`

```{r t-test exercise 1}
#t_test_results$

```

Now, try the t-test we did above, but try making the sample size much larger (say, n=1000 or n=10000). Do you notice a difference in the result of a t-test?

```{r t-test exercise 2}
#set.seed(20221011) # Seed for reproducibility
#n <-  # Sample size (CHANGE THIS)

#mu1 <- 0 # Mean of sample 1
#mu2 <- 0.1 # Mean of sample 2

#sd1 <- 1 # Standard deviation 1
#sd2 <- 1 # Standard deviation 2

#ve <- sd1 == sd2

# Generate two samples of of size n of normally-distributed data.
# Sample x1 has a mean of mu1 and standard deviation sd1
# Sample x2 has a mean of mu2 and standard deviation sd2
#x1 <- rnorm(n, mu1, sd1)
#x2 <- rnorm(n, mu2, sd2)

#t.test(x1,x2,var.equal = ve)
```

Try the t-test again. Make the sample size smaller (say, n=20) but change the means

```{r t-test exercise 3}
#set.seed(20221011) # Seed for reproducibility
#n <- 20 # Sample size

#mu1 <-  # Mean of sample 1 (CHANGE THIS)
#mu2 <-  # Mean of sample 2 (CHANGE THIS)

#sd1 <- 1 # Standard deviation 1
#sd2 <- 1 # Standard deviation 2

#ve <- sd1 == sd2

# Generate two samples of of size n of normally-distributed data.
# Sample x1 has a mean of mu1 and standard deviation sd1
# Sample x2 has a mean of mu2 and standard deviation sd2
#x1 <- rnorm(n, mu1, sd1)
#x2 <- rnorm(n, mu2, sd2)

#t.test(x1, x2, var.equal = ve)
```

### ANOVA and the F-test

What if the data have more than two groups? This leads naturally into Analysis of Variance (ANOVA).

We will create three "populations" with a measurement between them. These three groups will have different means to ensure that they give a statistically significant result.

```{r Create three artificial populations}
set.seed(20221011)

# Our artificial data will have three different means
m1 <- rnorm(30, mean = 25, sd = 5)
m2 <- rnorm(30, mean = 30, sd = 6)
m3 <- rnorm(30, mean = 22, sd = 7)

ms <- c(m1, m2, m3) # Combine our three measures into one vector

# Our vector of group labels A, B, and C
ls <- as.factor(c(rep("A",30),rep("B",30),rep("C",30)))

# Make it into a data frame
df_anova <- data.frame(ms, ls)
colnames(df_anova) <- c("measure","group")
```

First, we should visualize our data. We can use a boxplot to compare all the populations against each other.

```{r Visualize the three populations}
boxplot(measure ~ group, data = df_anova)
```

These groups certainly look different, but we should run an ANOVA just in case. We'll use the `aov` command. The command takes the formula `y ~ x` as an input, where `y` is the variable we are measuring and `x` is the grouping.

```{r Run the ANOVA}
# Run the ANOVA
results_aov <- aov(measure ~ group, data = df_anova)

# Call the results of the ANOVA
print("---Printing results of ANOVA---")
results_aov

# Print the summary
print("---Printing summary---")
summary(results_aov)
```

We now have a significant result. What else can we say? Let's use `$` to examine the ANOVA object:

```{r Look closer at the ANOVA}
results_aov$coefficients
```

These coefficients show us what our results are relative to Group A.

We can also use `plot` to see what our residuals look like.

```{r Plotting residuals of an ANOVA model}
plot(results_aov)
```

```         
(Intercept)      groupB      groupC 
  25.382285    4.521143   -4.900450 
```

### ANOVA exercises

Load the base R data `PlantGroup` using the command `data("PlantGrowth")`. Visualize the groups using a boxplot and run an ANOVA on them. Extract the coefficients using `$`.

```{r ANOVA exercise 1}
#data("PlantGrowth") # Load the data (it may take a second)

# inspect the first few rows

# create a box plot
```

```{r}
#run the ANOVA analysis

#check the outputs


#generate a summary

```

Run the `TukeyHSD()` command on the ANOVA you just created. Where are the differences in the model?
do pairwise comparisons between group means

```{r ANOVA exercise 2}

```

For two- and three-way ANOVA models, there are good examples at: <https://www.datanovia.com/en/lessons/anova-in-r/>. They require the `datarium` package.
